---
title: "Code implementing the Statistical Analysis Plans: Identifying biomarkers for cancer prognosis"
author:
  - name: Daniela Dunkler, Theresa Ullmann and Georg Heinze
    affiliation: Medical University of Vienna, Austria
  - name: Franziska Kappenberg
    affiliation: University of Bonn, Germany
  
date: last-modified
categories: [2026, statistical analysis plan, univariable selection, prediction, biomarker, case study, pre-defined model]
---

This document implements [the two statistical analysis
plans](web_extra_w2_study1_SAP.html), on the same 248-patient dataset,
comparing two contrasting analysis strategies for predicting disease-free survival
in breast cancer patients from PMP22 gene expression and four
established clinical factors (tumour size `pT`, nodal status `pN`,
differentiation grade `G`, and estrogen receptor status `ER`).

# Content {.unnumbered}

Two contrasting analyses for the same prediction problem are presented.

- @sec-analysis_bpa: Implementation of the pre-specified multivariable strategy consistent with current best-practice guidance (SAP A).
- @sec-analysis_uni: Implementation of the common but flawed workflow built on univariable variable selection (SAP B).
- @sec-analysis_contrast: Comparison of both approaches.

Both parts use *identical* data. Any difference in conclusions is
attributable to the **statistical strategy**, not to the data. The
underlying study is @tong2010gene. We work with 248 of the 249 patients
reported there [@pmp22data].

------------------------------------------------------------------------


# Data preparation (shared by both parts) {.unnumbered}

## Required packages {.unnumbered}

```{r}
#| label: setup
#| warning: false

library(survival)        # Cox regression, Surv objects
library(mice)            # multiple imputation by chained equations
library(rms)             # Harrell's regression modeling strategies toolkit
library(riskRegression)  # time-dependent AUC, Brier score, Index of Prediction Accuracy, bootstrap CV
library(MASS)            # stepAIC (used deliberately in the flawed Plan B)
library(car)             # generalized variance inflation factors
library(vcd)             # Cramer's V for the predictor-association heatmap (A.2)
library(DiagrammeR)      # patient flow chart (A.2)
library(ggplot2)
library(survminer)
library(knitr)

knitr::opts_chunk$set(fig.align = "center")

```

## Import and variable preparation {.unnumbered}

The data are loaded and prepared for analysis.

```{r}
#| label: load-data

df <- read.csv("data/data_pmp22.csv")
df$X <- NULL   # drop the row-index column

# PMP22, pN, ER are binary; pT and G are ordinal with 3 categories.
# We keep both a "raw" numeric/integer version (needed for linear/ordinal
# coding in Cox models) and labelled factor versions (needed for readable
# tables and plots).
df$PMP22_f <- factor(df$PMP22, levels = c(0, 1), labels = c("low", "high"))
df$pN_f    <- factor(df$pN,    levels = c(0, 1), labels = c("node-negative", "node-positive"))
df$ER_f    <- factor(df$ER,    levels = c(0, 1), labels = c("negative", "positive"))
df$pT_f    <- factor(df$pT,    levels = c(1, 2, 3), labels = c("pT1", "pT2", "pT3/pT4"), ordered = TRUE)
df$G_f     <- factor(df$G,     levels = c(1, 2, 3), labels = c("G1", "G2", "G3"), ordered = TRUE)

# str(df[, c("PMP22", "pT", "pN", "G", "ER", "days", "rezs")])

head(df)

```

------------------------------------------------------------------------

# SAP A: Pre-specified multivariable prediction model {#sec-analysis_bpa}

## A.1 Objective {.unnumbered}

A multivariable Cox model with all five pre-specified predictors
(`PMP22`, `pT`, `pN`, `G`, `ER`) entered simultaneously is estimated to
quantify the adjusted prognostic contribution of PMP22 alongside the
established clinical factors, and to report its prediction performance.
No predictor is excluded based on its own (univariable) association
with the outcome.

## A.2 Initial data analysis (IDA) {.unnumbered}

### Patient flow {.unnumbered}

Of the 249 patients reported in the original publication [@tong2010gene],
248 are available in the public dataset [@pmp22data] and form the
analysis population for both SAPs. Within these 248, 208 (83.9%) are
complete on all five predictors and enter the complete-case analyses
below. The remaining 40 (16.1%) have at least one missing predictor and
are handled via multiple imputation (primary analysis) or
excluded (complete-case sensitivity check).

```{r}
#| label: patient-flow
#| fig-cap: "Patient flow from the original publication to the analysis populations."

grViz('
digraph patient_flow {
  graph [layout = dot, rankdir = TB, nodesep = 0.4, ranksep = 0.45,
         fontsize = 11, fontname = Helvetica]
  node [shape = box, style = "rounded,filled", fillcolor = "#F2F2F2",
        color = "#4D4D4D", fontname = Helvetica, fontsize = 11,
        margin = "0.25,0.15"]
  edge [fontname = Helvetica, fontsize = 9, color = "#4D4D4D"]

  orig [label = "249 patients\\n(original publication)"]
  pub  [label = "248 patients\\n(public dataset;\\nanalysis population for both SAPs)"]
  cc   [label = "208 complete cases\\n(all 5 predictors observed)\\n=> complete-case analyses\\n(SAP B; SAP A sensitivity check)"]
  inc  [label = "40 patients with >=1\\nmissing predictor"]
  mi   [label = "All 248 patients,\\nmultiply imputed (m = 20)\\n=> SAP A primary analysis",
        fillcolor = "#E8F0FE"]

  orig -> pub [label = "1 patient not in\\npublic dataset"]
  pub  -> cc
  pub  -> inc
  pub  -> mi [style = dashed]
}
')

```

### Missingness {.unnumbered}

```{r}
#| label: ida-missing

miss_tab <- data.frame(
  variable = c("PMP22", "pT", "pN", "G", "ER"),
  n_missing = sapply(df[, c("PMP22", "pT", "pN", "G", "ER")], function(x) sum(is.na(x))),
  pct_missing = round(100 * sapply(df[, c("PMP22", "pT", "pN", "G", "ER")],
                                    function(x) mean(is.na(x))), 1)
)
kable(miss_tab, row.names = FALSE,
      caption = "Missing values per predictor (N = 248).")

n_complete <- sum(complete.cases(df[, c("PMP22", "pT", "pN", "G", "ER")]))
cat(sprintf("Complete cases on all 5 predictors: %d / %d (%.1f%%)\n",
            n_complete, nrow(df), 100 * n_complete / nrow(df)))

```

### Baseline characteristics {.unnumbered}

Descriptive distribution of the five predictors, complete cases only
(n = `r n_complete`). It is shown for plausibility-checking (A.2), not for
variable selection. No predictor is included or excluded on the basis
of these or any other univariate/outcome association (see A.6).

```{r}
#| label: baseline-table
#| tbl-cap: "Baseline characteristics of the complete-case analysis set."

cc0 <- df[complete.cases(df[, c("PMP22", "pT", "pN", "G", "ER")]), ]

baseline_row <- function(varname, label) {
  tab <- table(cc0[[varname]])
  data.frame(
    Variable = c(label, rep("", length(tab) - 1)),
    Category = names(tab),
    `N (%)`  = sprintf("%d (%.1f%%)", as.integer(tab), 100 * as.integer(tab) / sum(tab)),
    check.names = FALSE
  )
}

baseline_tab <- rbind(
  baseline_row("PMP22_f", "PMP22"),
  baseline_row("pN_f",    "Nodal status (pN)"),
  baseline_row("ER_f",    "Estrogen receptor (ER)"),
  baseline_row("pT_f",    "Tumour stage (pT)"),
  baseline_row("G_f",     "Grade (G)")
)
kable(baseline_tab, row.names = FALSE)

```

PMP22 and ER are strongly positively associated (chi-squared test, p < 0.001), which is
the correlation that masks PMP22's adjusted effect in the
univariable screen of SAP B (Section B.1):

```{r}
#| label: pmp22-er-crosstab

# helper: cross-tabulation with counts and row percentages, "n (row %)",
# plus a row total column; rows = x (percentages sum to 100% within each row)
xtab_rowpct <- function(x, y, xlab, ylab) {
  tab <- table(x, y)
  rs  <- rowSums(tab)
  out <- matrix(sprintf("%d (%.1f%%)", as.integer(tab), 100 * as.integer(tab) / rep(rs, ncol(tab))),
                nrow = nrow(tab), dimnames = list(NULL, colnames(tab)))
  out <- data.frame(rownames(tab), out, sprintf("%d (100%%)", as.integer(rs)),
                    check.names = FALSE)
  names(out) <- c(xlab, paste0(ylab, ": ", colnames(tab)), "Total")
  out
}

kable(xtab_rowpct(cc0$PMP22_f, cc0$ER_f, "PMP22", "ER"),
      row.names = FALSE,
      caption = "PMP22 by ER status, complete cases (n = 208); n (row %).")

```

As planned (A.2), the remaining pairwise cross-tabulations among the
five predictors are inspected as well, for plausibility only (not
against the outcome, and not for variable selection):

::: {.callout-tip collapse="true" title="Pairwise cross-tabulations among the five predictors"}

```{r}
#| label: predictor-crosstabs-all
#| results: asis

pred_vars   <- c("PMP22_f", "pN_f", "ER_f", "pT_f", "G_f")
pred_labels <- c(PMP22_f = "PMP22", pN_f = "Nodal status (pN)",
                  ER_f = "Estrogen receptor (ER)", pT_f = "Tumour stage (pT)",
                  G_f = "Grade (G)")

pred_pairs <- combn(pred_vars, 2, simplify = FALSE)
# the PMP22 x ER pair is shown separately above; the remaining 9 pairs follow
pred_pairs <- pred_pairs[!sapply(pred_pairs, function(p) all(c("PMP22_f","ER_f") %in% p))]

for (p in pred_pairs) {
  tab <- xtab_rowpct(cc0[[p[1]]], cc0[[p[2]]], pred_labels[p[1]], pred_labels[p[2]])
  cat("\n\n", pred_labels[p[1]], "by", pred_labels[p[2]], "; n (row %)\n\n")
  print(kable(tab, row.names = FALSE))
}

```

:::

As a compact complement to the `r choose(5,2)` individual
cross-tabulations, pairwise association strength is also summarised
visually using Cramer's V, which ranges from 0 (no association) to 1
(perfect association) and is a single-number summary for
cross-tabulations of categorical variables <!--with possibly different
numbers of levels-->:

```{r}
#| label: predictor-association-heatmap
#| fig-cap: "Pairwise association strength (Cramer's V) among the five predictors, complete cases (n = 208). The diagonal is fixed at 1 by definition."

assoc_mat <- matrix(1, nrow = length(pred_vars), ncol = length(pred_vars),
                     dimnames = list(pred_labels[pred_vars], pred_labels[pred_vars]))
for (p in combn(pred_vars, 2, simplify = FALSE)) {
  v <- suppressWarnings(assocstats(table(cc0[[p[1]]], cc0[[p[2]]]))$cramer)
  assoc_mat[pred_labels[p[1]], pred_labels[p[2]]] <- v
  assoc_mat[pred_labels[p[2]], pred_labels[p[1]]] <- v
}

assoc_df <- as.data.frame(as.table(assoc_mat))
names(assoc_df) <- c("Var1", "Var2", "CramersV")

ggplot(assoc_df, aes(Var1, Var2, fill = CramersV)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", CramersV)), size = 3.5) +
  scale_fill_gradient(low = "white", high = "#D7301F", limits = c(0, 1),
                       name = "Cramer's V") +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

The heatmap confirms the PMP22-ER association noted above (Cramer's V =
`r round(assoc_mat["PMP22","Estrogen receptor (ER)"],2)`) and shows no
other pairwise association among the five predictors strong enough to
raise plausibility concerns. As with the individual cross-tabulations,
this is a descriptive plausibility check (A.2).<!-- and plays no role in
variable selection.-->

`r n_complete` of 248 patients (`r round(100*n_complete/248,1)`%) are
complete on all five predictors. This moderate degree of missingness (up
to \~12% for nodal status) motivates using multiple imputation as the
primary analysis (Section A.5) rather than discarding \~16% of patients
via complete-case analysis.

### Plausibility of missing-at-random {.unnumbered}

As a simple, transparent check, we regress the missingness indicator of
each partially missing predictor on the other observed predictors and
the outcome (event indicator, log follow-up time). Significant
associations would suggest the missingness is related to observed
information, consistent with (but not proof of)
missing-at-random (MAR), and support using multiple imputation rather than
complete-case analysis.

```{r}
#| label: ida-mar-check
#| warning: FALSE

check_missing_assoc <- function(varname) {
  d <- df
  d$miss_ind <- as.integer(is.na(d[[varname]]))
  predictors <- setdiff(c("PMP22", "pT", "pN", "G", "ER"), varname)
  form <- as.formula(paste("miss_ind ~", paste(predictors, collapse = " + "),
                            "+ rezs + log(days)"))
  fit <- glm(form, data = d, family = binomial)
  round(summary(fit)$coefficients[, "Pr(>|z|)"], 3)
}

for (v in c("pT", "pN", "G")) {
  cat("\nMissingness of", v, "regressed on remaining variables + outcome:\n")
  print(check_missing_assoc(v))
}

```

None of these associations is strong or systematic, i.e., there is no
clear evidence *against* MAR. We proceed with multiple imputation as the
primary approach and report complete-case analysis as a sensitivity check.

*Note on ER completeness*: the original publication reports ER missing
in 22.5% of cases at the immunohistochemistry stage but re-classified
nearly all patients using an mRNA-expression-based surrogate. This
explains why ER is almost fully observed here (1 missing) despite the
more substantial original missingness. This is a data-provenance detail worth
documenting, not a statistical result.

## A.3 Outcome description {.unnumbered}

```{r}
#| label: ida-outcome
#| fig-cap: "Kaplan-Meier estimate of disease-free survival, all 248 patients."
#| message: FALSE

cat("Number of events (recurrence):", sum(df$rezs), "/", nrow(df),
    sprintf("(%.1f%%)\n", 100*mean(df$rezs)))

# Median potential follow-up via the reverse Kaplan-Meier method
# (Schemper & Smith, 1996) -- the appropriate way to summarise follow-up
# duration in survival studies, rather than simply the median observed time.
revkm <- survfit(Surv(days, 1 - rezs) ~ 1, data = df)
cat("Median potential follow-up (reverse KM):", summary(revkm)$table["median"],
    "days (~", round(summary(revkm)$table["median"] / 365.25, 1), "years)\n")

fit_km <- survfit(Surv(days, rezs) ~ 1, data = df)
ggsurvplot(fit_km, data = df, risk.table = TRUE, conf.int = TRUE,
           xlab = "Days since diagnosis", ylab = "Disease-free survival probability",
           legend = "none", palette = "black")

```

With 109 events among 248 patients (44.0%), and 5 predictors entered as
main effects (see A.4 below for their coding), events per parameter is
(`r round(109/5,1)` using linear/ordinal coding for `pT`/`G`,
`r round(109/7,1)` if coded as free 3-level factors instead).

## A.4 Functional form of ordinal predictors {.unnumbered}

`pT` and `G` are ordinal with 3 levels. We compare a parsimonious linear
(ordinal-score) coding against a free 3-level categorical coding via a
likelihood-ratio test, on the complete-case subset. (This functional-form
check is separate from, and does not constitute, variable *selection*, as 
all five predictors remain in the model regardless of the outcome.)

```{r}
#| label: functional-form

cc <- df[complete.cases(df[, c("PMP22", "pT", "pN", "G", "ER")]), ]
cat("Complete-case analysis set: n =", nrow(cc), "\n")

m_pT_lin <- coxph(Surv(days, rezs) ~ pT, data = cc)
m_pT_cat <- coxph(Surv(days, rezs) ~ factor(pT), data = cc)
cat("pT: linear vs. categorical coding, likelihood-ratio test p =",
    round(anova(m_pT_lin, m_pT_cat)$`Pr(>|Chi|)`[2], 3), "\n")

m_G_lin <- coxph(Surv(days, rezs) ~ G, data = cc)
m_G_cat <- coxph(Surv(days, rezs) ~ factor(G), data = cc)
cat("G:  linear vs. categorical coding, likelihood-ratio test p =",
    round(anova(m_G_lin, m_G_cat)$`Pr(>|Chi|)`[2], 3), "\n")

```

Neither test supports departure from linearity: both likelihood-ratio test p-values are above the pre-specified $\alpha = 0.05$ threshold (A.4). Hence, we retain
the more parsimonious linear (ordinal-score) coding for `pT` and `G`,
improving events-per-parameter without discarding information on their
ordering.

## A.6-A.8: Full model, assumption checks and interactions (complete-case) {.unnumbered}

Following the pre-specified analysis sequence (A.6): we first fit the
full main-effects model on complete cases and check multicollinearity
and proportional hazards (A.8) on it. Only then do we test the
pre-specified interactions (A.7), against the resulting (possibly
amended) model. This complete-case model is a diagnostic staging step. 
The final estimates come from the multiple-imputation model in Section
A.5 below.

```{r}
#| label: main-model-cc

fit_main <- coxph(Surv(days, rezs) ~ PMP22_f + pT + pN_f + G + ER_f,
                   data = cc, x = TRUE)
summary(fit_main)

```

### Multicollinearity (A.6/A.8) {.unnumbered}

```{r}
#| label: vif
#| warning: FALSE

vif(fit_main)

```

All generalized VIFs are close to 1. They are well below common concern
thresholds of 5 to 10. There is no problematic collinearity among the
five predictors. <!-- (`car::vif()` warns "No intercept: vifs may not be
sensible" — this is expected and harmless for a `coxph` fit, which by
construction never has an intercept term (it is absorbed into the
baseline hazard); the warning is not evidence of a coding error here.) -->

### Proportional hazards assumption (A.8) {.unnumbered}

```{r}
#| label: ph-check
#| fig-cap: "Scaled Schoenfeld residuals vs. time for each covariate."
#| layout-ncol: 2

zph <- cox.zph(fit_main)
zph
plot(zph)

```

The proportional hazards assumption is clearly violated for `ER_f` (p \<
0.001), with the global test also significant. This mirrors what @tong2010gene themselves report and address in the source publication. We adopt
the same remedy: a time-varying effect of ER, modelled as an interaction
between ER and the logarithm of survival time. In code this is written as `ER_f x log(time)`. We refit the model with a time-varying effect of ER. 

<!--`PMP22_f` also shows a borderline-significant per-covariate test (p
$\approx$ 0.026), on its own smaller than ER's but formally below 0.05.
It is not acted upon here (only the ER violation, which is far
stronger, is remedied). We note it explicitly rather than silently pass
over it, since it is a residual caveat on the proportional-hazards
assumption for the variable of primary interest.-->

```{r}
#| label: time-varying-ER

fit_main_tv <- coxph(
  Surv(days, rezs) ~ PMP22_f + pT + pN_f + G + ER_f + tt(ER_f),
  data = cc,
  tt = function(x, t, ...) (as.numeric(x) - 1) * log(t)
)
summary(fit_main_tv)

```

**Interpreting the time-varying effect of ER:** in the model above the time-varying effect of ER is represented with `tt(ER_f)`. With this coding, the
model estimates the ER-positive-vs-negative log hazard ratio at
follow-up time $t$ (days) as $\beta_{ER} + \beta_{ER\times\log(t)}
\times \log(t)$. The coefficient printed for `ER_f` is the
log hazard ratio at $t = 1$ day (where $\log(t) = 0$), not an
average, time-constant effect. 

The coefficient printed for `tt(ER_f)` is the change in that log hazard ratio per unit increase in
$\log(t)$, i.e., it quantifies *how* the ER effect evolves over
follow-up rather than a hazard ratio in its own right. So it should
not itself be exponentiated and reported as a hazard ratio. Here the
estimated `tt(ER_f)` coefficient is
`r round(coef(fit_main_tv)[["tt(ER_f)"]],3)`<!--, which is
`r ifelse(coef(fit_main_tv)[["tt(ER_f)"]] < 0, "negative", "positive")`:
the ER effect
`r ifelse(coef(fit_main_tv)[["tt(ER_f)"]] < 0,
"attenuates (weakens, and beyond some follow-up time may reverse) the longer patients are followed -- consistent with the well-documented late loss of the early ER-positive prognostic advantage in breast cancer",
"strengthens the longer patients are followed")`-->. A positive coefficient means the ER effect shifts towards higher hazard (i.e., the protective effect of ER-positivity
weakens) with longer follow-up, while a negative coefficient would mean the opposite. 

The ER hazard ratio at any specific follow-up time can be obtained by
evaluating $\exp(\beta_{ER} + \beta_{ER\times\log(t)}\times\log(t))$ at
that value of $t$.

Accounting for the time-varying ER effect materially changes the
estimated PMP22 effect: on complete cases, the PMP22 hazard ratio moves
from `r round(exp(coef(fit_main))["PMP22_fhigh"],2)` (95% CI
`r round(summary(fit_main)$conf.int["PMP22_fhigh","lower .95"],2)`-`r round(summary(fit_main)$conf.int["PMP22_fhigh","upper .95"],2)`;
p = `r format.pval(summary(fit_main)$coefficients["PMP22_fhigh","Pr(>|z|)"], digits = 3, eps = 0.001)`)
in the standard model to
`r round(exp(coef(fit_main_tv))["PMP22_fhigh"],2)` (95% CI
`r round(summary(fit_main_tv)$conf.int["PMP22_fhigh","lower .95"],2)`-`r round(summary(fit_main_tv)$conf.int["PMP22_fhigh","upper .95"],2)`;
p = `r format.pval(summary(fit_main_tv)$coefficients["PMP22_fhigh","Pr(>|z|)"], digits = 3, eps = 0.001)`)
once the ER proportional-hazards violation is corrected. **This illustrates
that assumption-checking is not a formality.** Correcting the ER term 
also changes what we conclude about PMP22, the variable of primary
interest. <!-- (`cox.zph()` does not support models with `tt()` terms
directly; a supplementary check using explicit time-splitting, not shown
in full here, indicates that any residual time-dependence in the
remaining covariates is mild.) -->

### Interactions (A.7), tested against the assumption-corrected model {.unnumbered}

The plan's pre-specified analysis sequence (A.6) requires that
interactions be tested only *after* proportional-hazards/functional-form
assumptions have been checked, and against the resulting (possibly
amended) main-effects model (i.e., the model with the `ER`
time-varying term) rather than against the original, proportional hazards-violating
specification. A small, pre-specified set of clinically motivated
interactions (`PMP22 x ER`, `PMP22 x pN`, `PMP22 x pT`) is tested
jointly with a single likelihood-ratio test, to avoid inflating the
false-positive rate that would result from testing every pairwise
combination individually.

```{r}
#| label: interactions

fit_int_tv <- coxph(
  Surv(days, rezs) ~ PMP22_f * ER_f + PMP22_f * pN_f + PMP22_f * pT + G + tt(ER_f),
  data = cc,
  tt = function(x, t, ...) (as.numeric(x) - 1) * log(t)
)
int_test  <- anova(fit_main_tv, fit_int_tv)
p_int_cc  <- int_test[["Pr(>|Chi|)"]][2]
int_test

```

At the pre-specified $\alpha = 0.05$ (A.7), the joint test is
`r ifelse(is.na(p_int_cc), "inconclusive", ifelse(p_int_cc < 0.05, "significant", "not significant"))`
(p = `r round(p_int_cc, 3)`). Per A.7, unless an interaction is both
statistically compelling and clinically sensible, the primary reported
prediction model retains only main effects (plus the `ER` time-varying
term). Any retained interaction would in any case be treated as
exploratory/hypothesis-generating given the limited events-per-parameter.

## A.5 Multiple imputation and pooled final model {.unnumbered}

We now impute the (moderate) missingness in the five predictors using
multiple imputation by chained equations, and fit the model settled on
above (linear `pT`/`G`, time-varying `ER`, no interactions) on each
imputed dataset, pooling with Rubin's rules.

Following @white_2009, the imputation model for the
predictors includes the **Nelson–Aalen estimator of the cumulative
hazard** and the **event indicator**, rather than raw survival time, to
preserve the covariate–outcome associations correctly.

```{r}
#| label: multiple-imputation

set.seed(852945)

df_mi <- df[, c("PMP22", "pT", "pN", "G", "ER", "days", "rezs")]
df_mi$nelsonH <- nelsonaalen(df_mi, days, rezs)
df_mi$PMP22 <- factor(df_mi$PMP22)
df_mi$pN    <- factor(df_mi$pN)
df_mi$ER    <- factor(df_mi$ER)
df_mi$pT    <- factor(df_mi$pT, ordered = TRUE)
df_mi$G     <- factor(df_mi$G,  ordered = TRUE)

imp0 <- mice(df_mi, maxit = 0)
meth <- imp0$method
meth["PMP22"] <- "logreg"; meth["pN"] <- "logreg"; meth["ER"] <- "logreg"
meth["pT"] <- "polr";      meth["G"]  <- "polr"
meth["days"] <- ""; meth["rezs"] <- ""; meth["nelsonH"] <- ""

pred <- imp0$predictorMatrix
pred[, "days"] <- 0   # use nelsonH + rezs instead of raw survival time
pred["days", ] <- 0
diag(pred) <- 0

# m chosen following White, Royston & Wood's rule of thumb:
# m >= 100 * fraction of incomplete cases (~16% here -> m >= 16); we use 20.
imp <- mice(df_mi, method = meth, predictorMatrix = pred, m = 20, maxit = 20,
            printFlag = FALSE, seed = 2026)

fit_imp <- with(imp, coxph(
  Surv(days, rezs) ~ PMP22 + as.numeric(pT) + pN + as.numeric(G) + ER + tt(ER),
  tt = function(x, t, ...) (as.numeric(x) - 1) * log(t)
))
pooled <- pool(fit_imp)
s <- summary(pooled, conf.int = TRUE)

hr_terms <- c("PMP22 (high vs. low)", "pT (per category)", "pN (positive vs. negative)",
              "G (per grade)", "ER (positive vs. negative, at t=1 day)", "ER x log(time)")

# 2 decimals throughout, except the ER HR at t = 1 day: it is close to 0 and
# would be reported as 0.00 with 2 decimals, so it is given to 4 decimals.
hr_digits <- ifelse(hr_terms == "ER (positive vs. negative, at t=1 day)", 4, 2)
fmt_hr <- function(x, d) unname(mapply(function(xx, dd) formatC(xx, format = "f", digits = dd),
                                       x, d))

hr_raw <- exp(s$estimate)   # unrounded HRs, used for derived quantities below

hr_table <- data.frame(
  term    = hr_terms,
  HR      = round(hr_raw, hr_digits),
  CI_low  = round(exp(s$`2.5 %`), hr_digits),
  CI_high = round(exp(s$`97.5 %`), hr_digits),
  p_value = s$p.value   # kept at full precision for re-use in the text below
)

hr_display <- hr_table
hr_display$HR      <- fmt_hr(hr_table$HR,      hr_digits)
hr_display$CI_low  <- fmt_hr(hr_table$CI_low,  hr_digits)
hr_display$CI_high <- fmt_hr(hr_table$CI_high, hr_digits)
hr_display$p_value <- format.pval(hr_display$p_value, digits = 3, eps = 0.001)

# the ER x log(time) term is not a hazard ratio: show only its p-value
# (HR and CI blank); the coefficient is given in the footnote below
i_tt <- hr_display$term == "ER x log(time)"
hr_display$HR[i_tt] <- ""; hr_display$CI_low[i_tt] <- ""; hr_display$CI_high[i_tt] <- ""
hr_display$term[i_tt] <- "ER x log(time)^a^"

# pooled coefficient (log scale) and SE of the time-varying ER term, for the footnote
b_tt  <- s$estimate[hr_terms == "ER x log(time)"]
se_tt <- s$std.error[hr_terms == "ER x log(time)"]

kable(hr_display, row.names = FALSE,
      caption = "Final pooled multivariable Cox model (m = 20 imputations, Rubin's rules).")

```

^a^ Time-varying effect of ER, modelled as ER x log(time): the pooled coefficient is 
$\hat\beta_{ER \times \log(t)}$ = `r sprintf("%.3f", b_tt)` (SE `r sprintf("%.3f", se_tt)`) on the log-hazard scale.
It is the change in the ER log hazard ratio per one-unit increase in log(days)<!--, so the ratio of ER hazard ratios
per one-unit increase in log(days) is $\exp(\hat\beta)$ = `r sprintf("%.2f", exp(b_tt))`.
The ER log hazard ratio at time $t$ (days) is $\hat\beta_{ER} + \hat\beta_{ER \times \log(t)} \log(t)$,
where $\hat\beta_{ER}$ is the ER coefficient at $t$ = 1 day (the row "ER (positive vs. negative, at t=1 day)").
A positive coefficient means the ER effect shifts towards higher hazard (the protective effect of ER-positivity
weakens) with longer follow-up; a negative coefficient means the opposite.-->


### Congenial interaction test under multiple imputation {.unnumbered}

As noted in the plan (A.5), testing the pre-specified interactions only
on complete cases (as done above, in A.6/A.7) is not congenial with an
imputation model that itself excludes those interaction terms. Because
`PMP22`, `ER`, `pN`, and `pT` are imputed directly, the
interaction products can be formed on the fly, within `with()`, from
the already-imputed variables in each of the 20 completed datasets. This needs no separate passive-imputation step. We fit an 
interaction-augmented version of the pooled model and compare it to
`fit_imp` with a multivariate Wald test (`mice::D1()`), the standard
multiple-imputation analogue of a joint likelihood-ratio test.

```{r}
#| label: mi-interactions

fit_imp_int <- with(imp, coxph(
  Surv(days, rezs) ~ PMP22 * ER + PMP22 * pN + PMP22 * as.numeric(pT) +
    as.numeric(pT) + as.numeric(G) + tt(ER),
  tt = function(x, t, ...) (as.numeric(x) - 1) * log(t)
))
D1(fit_imp_int, fit_imp)

```

This multiple imputation-based joint test is the primary check for the interaction terms.
The complete-case version above (A.6/A.7) is retained as a consistency
check. If both agree that the interactions are not significant at the
pre-specified $\alpha = 0.05$, the main-effects-only (plus `ER`
time-varying) model above remains the primary reported result. If they
disagree, this would need to be reconciled before finalizing the model.


**Primary result:** *after multiple imputation and adjustment
for the four established clinical factors, patients with high PMP22
expression have a
`r hr_table$HR[hr_table$term=="PMP22 (high vs. low)"]`-fold (95% CI
`r hr_table$CI_low[hr_table$term=="PMP22 (high vs. low)"]` - `r hr_table$CI_high[hr_table$term=="PMP22 (high vs. low)"]`)
higher hazard of recurrence than patients with low PMP22 expression (p =
`r format.pval(hr_table$p_value[hr_table$term=="PMP22 (high vs. low)"], digits = 3, eps = 0.001)`).*
The point estimate is directionally consistent with, but numerically somewhat
lower than, the published estimate (HR 1.77, p = 0.02) and than the
complete-case estimate with the same time-varying-ER specification (HR
`r round(exp(coef(fit_main_tv))["PMP22_fhigh"],2)`, p =
`r round(summary(fit_main_tv)$coefficients["PMP22_fhigh","Pr(>|z|)"],3)`,
Section A.6-A.8 above). The wider confidence interval and the p-value no
longer crossing the conventional 0.05 threshold reflect that multiple
imputation propagates the extra uncertainty from the 16.1% of cases with
missing predictors, which the complete-case analysis cannot. The lower
point estimate may reflect the imputation itself (cases with missing
predictors contribute information that the complete-case analysis
discards) and the slightly different sample (248 vs. 249 patients). The
data alone cannot tell these apart. Taken together, both analyses agree
on the qualitative, substantive conclusion motivating this case study:
**PMP22's adjusted association with recurrence is materially stronger
than its univariable association** (HR 1.01, p = 0.98, Section B.1). This is 
the opposite of what a univariable screen would suggest (see SAP B), 
even though, after propagating imputation uncertainty, the adjusted
estimate itself is no longer significant at the conventional 5% level.

## A.9 Prediction performance and internal validation {.unnumbered}

### Multiple imputation-pooled bootstrap optimism-correction (primary analysis) {.unnumbered}

Per the plan (A.9), the bootstrap optimism-correction is combined with
multiple imputation by repeating the full bootstrap procedure *within*
each of the 20 imputed datasets separately<!--, rather than bootstrapping a
single stacked (completed) dataset-->. The resulting optimism-corrected
estimates from each imputed dataset are then pooled across imputations
using Rubin's rules, so that imputation uncertainty is reflected in the
reported standard error<!-- rather than being lost in the resampling step-->.

We illustrate this for Harrell's C-index, the calibration slope, the
explained-variation measure (Nagelkerke's pseudo-R^2 for Cox models),
the time-dependent AUC, Brier score and Index of Prediction Accuracy
summary, all using the *non-time-varying* model specification. Pooling proceeds the same way
throughout: the full bootstrap (or bootstrap-cross-validation) procedure
is repeated within each of the 20 imputed datasets, and the resulting
estimates are combined across imputations using Rubin's rules.

*Why the non-time-varying model is used for all performance measures.*
The C-index, the calibration slope and the explained-variation measure
summarise how well a single risk score per patient (the linear
predictor) ranks and calibrates patients. With a time-varying effect of
`ER`, the patients' ranking changes over follow-up, so no single risk
score exists. For the time-dependent AUC, Brier score and Index of
Prediction Accuracy we use the same model, so that all performance
measures describe the same model. <!--Repeating the full bootstrap
in each of the 20 imputed datasets is also much simpler and faster for
this model.-->

*Likely consequences.* The constant `ER` effect is an average over
follow-up, and the proportional hazards assumption is violated for `ER`
(A.8). Therefore, the model whose performance is reported here differs
from the final model in A.5 (Table 2) including the time-varying
`ER` term and is probably somewhat less accurate. Discrimination and
calibration may thus be slightly mis-estimated, in particular for
predictions early or late in follow-up. Because only one of the five
predictors is affected, we would expect the differences to be small. <!--,
but this was not checked here.--> The results should be read as
approximate.

```{r}
#| label: mi-bootstrap-cindex

# Bootstrap optimism-correction for Harrell's C-index, the calibration
# slope, and the explained-variation measure (Nagelkerke's pseudo-R^2 for
# Cox models), repeated within each imputed dataset and pooled across
# imputations with Rubin's rules. The calibration-slope and R^2
# replicates reuse the same bootstrap-fitted model as the C-index
# replicate (fit_b):
# - slope_test is the coefficient obtained from refitting Surv ~ lp on
#   the original data using fit_b's linear predictor (Steyerberg-style
#   calibration slope); a value < 1 indicates overfitting.
# - r2_test is Nagelkerke's R^2 (Nagelkerke, 1991), computed from the
#   partial log-likelihood of fit_b's linear predictor evaluated on the
#   original data (via a zero-free-parameter, offset-only Cox model),
#   relative to the null-model log-likelihood on the original data. This
#   is the same explained-variation quantity reported by rms::validate()
#   (row "R2"); it is rebuilt here by hand, rather than taken from
#   rms::validate(), so that the per-replicate values needed for
#   Rubin's-rules pooling are accessible.
nagelkerke_r2 <- function(loglik_model, loglik_null, n) {
  LR    <- 2 * (loglik_model - loglik_null)
  R2_cs <- 1 - exp(-LR / n)
  R2_cs / (1 - exp(2 * loglik_null / n))
}

boot_optimism_C <- function(data, B = 100) {
  form  <- Surv(days, rezs) ~ PMP22_f + pT_num + pN_f + G_num + ER_f
  fit0  <- coxph(form, data = data, x = TRUE)
  c_app <- concordance(fit0)$concordance

  n <- nrow(data)
  null_orig     <- coxph(Surv(days, rezs) ~ 1, data = data)
  loglik_null_o <- null_orig$loglik[1]
  r2_app        <- nagelkerke_r2(fit0$loglik[2], loglik_null_o, n)

  c_train <- numeric(B); c_test <- numeric(B); slope_test <- numeric(B)
  r2_train <- numeric(B); r2_test <- numeric(B)
  for (b in seq_len(B)) {
    boot_data <- data[sample(seq_len(n), n, replace = TRUE), ]
    fit_b <- tryCatch(coxph(form, data = boot_data, x = TRUE),
                       error = function(e) NULL)
    if (is.null(fit_b)) {
      c_train[b] <- NA; c_test[b] <- NA; slope_test[b] <- NA
      r2_train[b] <- NA; r2_test[b] <- NA
      next
    }
    c_train[b] <- concordance(fit_b)$concordance
    lp_orig   <- predict(fit_b, newdata = data, type = "lp")
    # NOTE: concordance()'s formula method assumes, by default, that LARGER
    # x means LONGER survival (opposite convention to a coxph risk score,
    # where larger lp means shorter survival) -- reverse = TRUE is required
    # here, or the resulting "concordance" is 1 minus the true value.
    c_test[b] <- concordance(Surv(data$days, data$rezs) ~ lp_orig,
                              reverse = TRUE)$concordance
    slope_fit <- tryCatch(coxph(Surv(data$days, data$rezs) ~ lp_orig),
                           error = function(e) NULL)
    slope_test[b] <- if (is.null(slope_fit)) NA else coef(slope_fit)[1]

    null_b <- tryCatch(coxph(Surv(days, rezs) ~ 1, data = boot_data),
                        error = function(e) NULL)
    r2_train[b] <- if (is.null(null_b)) NA else
      nagelkerke_r2(fit_b$loglik[2], null_b$loglik[1], n)
    test_fit <- tryCatch(coxph(Surv(data$days, data$rezs) ~ offset(lp_orig)),
                          error = function(e) NULL)
    r2_test[b] <- if (is.null(test_fit)) NA else
      nagelkerke_r2(test_fit$loglik[1], loglik_null_o, n)
  }
  optimism_C  <- mean(c_train - c_test, na.rm = TRUE)
  optimism_R2 <- mean(r2_train - r2_test, na.rm = TRUE)
  # The apparent calibration slope of a model against its own training data
  # is 1 by construction (lp already solves the partial-likelihood score
  # equations), so optimism = 1 - mean(slope_test) and the corrected slope
  # is simply mean(slope_test).
  list(apparent = c_app, corrected = c_app - optimism_C,
       boot_var = var(c_test, na.rm = TRUE),
       slope_corrected = mean(slope_test, na.rm = TRUE),
       slope_var = var(slope_test, na.rm = TRUE),
       R2_apparent = r2_app, R2_corrected = r2_app - optimism_R2,
       R2_var = var(r2_test, na.rm = TRUE))
}

set.seed(20260826)
m_imp <- imp$m
mi_c_results <- vector("list", m_imp)
for (i in seq_len(m_imp)) {
  d_i <- complete(imp, i)
  d_i$PMP22_f <- factor(as.character(d_i$PMP22), levels = c("0", "1"),
                         labels = c("low", "high"))
  d_i$pN_f    <- factor(as.character(d_i$pN), levels = c("0", "1"),
                         labels = c("node-negative", "node-positive"))
  d_i$ER_f    <- factor(as.character(d_i$ER), levels = c("0", "1"),
                         labels = c("negative", "positive"))
  d_i$pT_num  <- as.numeric(as.character(d_i$pT))
  d_i$G_num   <- as.numeric(as.character(d_i$G))
  mi_c_results[[i]] <- boot_optimism_C(d_i, B = 100)
}

corrected_vec <- sapply(mi_c_results, `[[`, "corrected")
apparent_vec  <- sapply(mi_c_results, `[[`, "apparent")
within_var    <- mean(sapply(mi_c_results, `[[`, "boot_var"))
between_var   <- var(corrected_vec)
pooled_C      <- mean(corrected_vec)
pooled_se     <- sqrt(within_var + (1 + 1 / m_imp) * between_var)

slope_vec        <- sapply(mi_c_results, `[[`, "slope_corrected")
slope_within_var <- mean(sapply(mi_c_results, `[[`, "slope_var"))
slope_between_var<- var(slope_vec)
pooled_slope     <- mean(slope_vec)
pooled_slope_se  <- sqrt(slope_within_var + (1 + 1 / m_imp) * slope_between_var)

r2_vec         <- sapply(mi_c_results, `[[`, "R2_corrected")
r2_within_var  <- mean(sapply(mi_c_results, `[[`, "R2_var"))
r2_between_var <- var(r2_vec)
pooled_R2      <- mean(r2_vec)
pooled_R2_se   <- sqrt(r2_within_var + (1 + 1 / m_imp) * r2_between_var)

cat(sprintf("Apparent C, averaged across imputations:                %.3f\n",
            mean(apparent_vec)))
cat(sprintf("Bootstrap optimism-corrected C, pooled (Rubin):         %.3f (SE %.3f)\n",
            pooled_C, pooled_se))
cat(sprintf("Bootstrap optimism-corrected calib. slope, pooled:      %.3f (SE %.3f)\n",
            pooled_slope, pooled_slope_se))
cat(sprintf("Bootstrap optimism-corrected explained variation\n  (Nagelkerke R2), pooled (Rubin):                       %.3f (SE %.3f)\n",
            pooled_R2, pooled_R2_se))

```

Apparent discrimination (fit to the same data used to estimate the
model) overstates true predictive ability. The pooled, optimism-corrected
C-index above is the more honest number to report. The optimism-corrected calibration slope
(`r round(pooled_slope,2)`, i.e., somewhat below the ideal value of 1)
indicates mild overfitting/optimism, plausible given the modest number of events
relative to the number of parameters. Shrinking new predictions by
this factor before use would be a reasonable. 
The pooled, optimism-corrected explained variation (`r round(pooled_R2,2)`,
Nagelkerke's pseudo-R^2^ for Cox models) summarises the proportion of
variation in disease-free survival accounted for by the model.<!-- It is
the multiple imputation-pooled counterpart to the "proportion of
explained variation" reported in the original publication, and to the
Index of Prediction Accuracy reported for the complete-case sensitivity
check below -- all three are conceptually related explained-variation
summaries, but are estimated differently and are not expected to be
numerically identical.-->

#### Time-dependent AUC, Brier score and Index of Prediction Accuracy (multiple imputation-pooled) {.unnumbered}

The time-dependent AUC, Brier score and Index of Prediction Accuracy
are pooled the same way: the bootstrap-cross-validated
`riskRegression::Score()` call is repeated within each of the 20
imputed datasets, and the resulting AUC(t) and Brier(t) estimates 
(together with their bootstrap-cross-validation standard errors) are
combined across imputations using Rubin's rules.

```{r}
#| label: mi-pooled-score
#| warning: false

set.seed(20260827)
mi_score_results <- vector("list", m_imp)
for (i in seq_len(m_imp)) {
  d_i <- complete(imp, i)
  d_i$PMP22_f <- factor(as.character(d_i$PMP22), levels = c("0", "1"),
                         labels = c("low", "high"))
  d_i$pN_f    <- factor(as.character(d_i$pN), levels = c("0", "1"),
                         labels = c("node-negative", "node-positive"))
  d_i$ER_f    <- factor(as.character(d_i$ER), levels = c("0", "1"),
                         labels = c("negative", "positive"))

  # standard (non-time-varying) model, as for the C-index, calibration slope and R^2
  fit_i <- coxph(Surv(days, rezs) ~ PMP22_f + pT + pN_f + G + ER_f,
                  data = d_i, x = TRUE, y = TRUE)

  sc_i <- Score(list("Full model" = fit_i),
                formula = Surv(days, rezs) ~ 1,
                data = d_i,
                times = c(365, 3 * 365, 5 * 365),
                metrics = c("auc", "brier"),
                summary = "ipa",
                split.method = "bootcv", B = 100, M = 150,
                progress.bar = NULL)

  auc_i   <- sc_i$AUC$score[sc_i$AUC$score$model   == "Full model", ]
  brier_i <- sc_i$Brier$score[sc_i$Brier$score$model == "Full model", ]

  mi_score_results[[i]] <- list(auc = auc_i$AUC, auc_se = auc_i$se,
                                 brier = brier_i$Brier, brier_se = brier_i$se,
                                 ipa = brier_i$IPA)
}

rubin_pool <- function(est_mat, se_mat) {
  # est_mat, se_mat: m (imputations) x k (time horizons) matrices
  m <- nrow(est_mat)
  pooled_est  <- colMeans(est_mat)
  within_var  <- colMeans(se_mat^2)
  between_var <- apply(est_mat, 2, var)
  data.frame(estimate = pooled_est,
             se = sqrt(within_var + (1 + 1 / m) * between_var))
}

auc_mat      <- do.call(rbind, lapply(mi_score_results, `[[`, "auc"))
auc_se_mat   <- do.call(rbind, lapply(mi_score_results, `[[`, "auc_se"))
brier_mat    <- do.call(rbind, lapply(mi_score_results, `[[`, "brier"))
brier_se_mat <- do.call(rbind, lapply(mi_score_results, `[[`, "brier_se"))
ipa_mat      <- do.call(rbind, lapply(mi_score_results, `[[`, "ipa"))

pooled_auc   <- rubin_pool(auc_mat, auc_se_mat)
pooled_brier <- rubin_pool(brier_mat, brier_se_mat)
# Score() does not return a per-replicate standard error for IPA, so a
# full Rubin's-rules combination (which needs a within-imputation
# variance) is not directly available for this metric. The pooled point
# estimate (mean across imputations) is reported together with the
# between-imputation standard deviation, as a simpler summary of
# stability across imputed datasets -- see the note below the tables.
pooled_ipa <- data.frame(estimate = colMeans(ipa_mat),
                          between_sd = apply(ipa_mat, 2, sd))

horizon <- c("1 year", "3 years", "5 years")

kable(cbind(Horizon = horizon, round(pooled_auc, 3)), row.names = FALSE,
      col.names = c("Horizon", "Pooled AUC(t)", "SE"),
      caption = "Multiple imputation-pooled, bootstrap-cross-validation-corrected time-dependent AUC(t) (m = 20 imputations, Rubin's rules).")

kable(cbind(Horizon = horizon, round(pooled_brier, 3)), row.names = FALSE,
      col.names = c("Horizon", "Pooled Brier score", "SE"),
      caption = "Multiple imputation-pooled, bootstrap-cross-validation-corrected Brier score (m = 20 imputations, Rubin's rules).")

kable(cbind(Horizon = horizon, round(pooled_ipa, 3)), row.names = FALSE,
      col.names = c("Horizon", "Pooled IPA", "Between-imputation SD"),
      caption = "Multiple imputation-pooled Index of Prediction Accuracy (IPA) (m = 20 imputations; point estimate and between-imputation SD -- see note below).")
```

The pooled AUC(t), Brier score and index of prediction accuracy estimates above are the primary,
multiple imputation-based counterparts of the complete-case sensitivity
check reported further below. As with the C-index, calibration slope
and explained variation, pooling here reflects imputation uncertainty in
the standard errors (where available) rather than discarding it in the
resampling step. For index of prediction accuracy specifically, note that the between-imputation
standard deviation (SD) reported above captures only variability *across*
imputed datasets, not the within-imputation sampling uncertainty that
Rubin's rules would normally add in. It should be read as a
lower bound on the true uncertainty of the pooled index of prediction accuracy estimate, unlike the AUC(t) and Brier score standard errors, which are full
Rubin's-rules combinations.

### Complete-case sensitivity check {.unnumbered}

As a simpler, secondary check we run the standard `rms` bootstrap
optimism-correction workflow once, on the complete-case data only,
**without** multiple imputation. This does not reflect imputation
uncertainty (so it is expected to differ somewhat from the pooled
multiple imputation-based figures above) and is reported only for
comparison. All three sub-analyses below (`rms::validate()`, the
calibration plot, and the time-dependent AUC, Brier score and Index of
Prediction Accuracy summary) use the same complete-case dataset `cc`
(n = `r nrow(cc)`). In
particular, the time-dependent AUC(t) analysis in this subsection is
the complete-case counterpart to the multiple imputation-pooled version
reported just above: the two use different data (208 complete cases
here vs. all 248 patients, via multiple imputation, above) and are
therefore expected to differ somewhat from one another. As in the primary
analysis, all complete-case performance measures refer to the standard
(non-time-varying) model with a constant `ER` effect.


#### rms::validate() bootstrap C-index (complete-case) {.unnumbered}

```{r}
#| label: rms-validation

dd <- datadist(cc); options(datadist = "dd")

fit_cph <- cph(Surv(days, rezs) ~ PMP22_f + pT + pN_f + G + ER_f, data = cc,
               x = TRUE, y = TRUE, surv = TRUE, time.inc = 365 * 5)
fit_cph

set.seed(2026)
val <- validate(fit_cph, method = "boot", B = 300)
val
```

```{r}
#| label: derived-c-index

Dxy_app <- val["Dxy", "index.orig"]
Dxy_cor <- val["Dxy", "index.corrected"]
cat(sprintf("Apparent Harrell's C:            %.3f\n", 0.5 + Dxy_app/2))
cat(sprintf("Bootstrap optimism-corrected C:  %.3f\n", 0.5 + Dxy_cor/2))
```

#### Calibration (complete-case) {.unnumbered}
 
```{r}
#| label: calibration
#| fig-cap: "Bootstrap-corrected calibration at 5 years (predicted vs. observed DFS probability), complete-case sensitivity check."

set.seed(2026)
cal <- calibrate(fit_cph, method = "boot", B = 300, u = 365 * 5, m = 50)
plot(cal, xlab = "Predicted 5-year DFS probability",
     ylab = "Observed 5-year DFS probability (bootstrap-corrected)")
```

<br>

*Note:* two different calibration metrics appear in this document. The calibration
**slope** reported earlier (Section "Multiple imputation-pooled
bootstrap optimism-correction") is pooled across the 20 imputed
datasets, consistent with multiple imputation as the primary analysis
strategy<!--; the same section also reports the multiple imputation-pooled
explained-variation measure and the multiple imputation-pooled
time-dependent AUC, Brier score and Index of Prediction Accuracy
summary-->. The calibration **plot** shown above, by contrast, belongs to
this complete-case sensitivity check.<!--: unlike the metrics just listed,
it is estimated on the complete-case dataset only and does not use the
imputed datasets. Extending the calibration *plot itself* to a full multiple
imputation + bootstrap loop -- i.e. averaging predicted-vs-observed
calibration curves across the 20 imputed datasets -- follows the
identical logic demonstrated above for the other metrics, but is
visually and computationally more involved (there is no single scalar
to pool with Rubin's rules; the whole curve would need to be combined)
and is left as a documented scope limitation.-->

#### Time-dependent AUC, Brier score and Index of Prediction Accuracy (complete-case) {.unnumbered}
 
This is the complete-case sensitivity-check counterpart to the multiple imputation-pooled version reported in Section "Multiple imputation-pooled bootstrap
optimism-correction". As a modern complement to the C-index/calibration-slope summary above, we use `riskRegression::Score()` with bootstrap cross-validation to
obtain time-dependent AUC(t), the Brier score, and the Index of
Prediction Accuracy (IPA<!--; conceptually analogous to the "proportion of
explained variation" reported in the original publication, though
estimated differently-->) at 1, 3 and 5 years. 

```{r}
#| label: score-performance
#| message: false

fit_std <- coxph(Surv(days, rezs) ~ PMP22_f + pT + pN_f + G + ER_f,
                  data = cc, x = TRUE, y = TRUE)

set.seed(202625)
sc <- Score(list("Full model" = fit_std),
            formula = Surv(days, rezs) ~ 1,
            data = cc,
            times = c(365, 3 * 365, 5 * 365),
            metrics = c("auc", "brier"),
            summary = "ipa",
            split.method = "bootcv", B = 100, M = 150,
            progress.bar = NULL)
sc

```

**Interpreting the output:** 
  - AUC(t) is Uno's time-dependent area under
the ROC curve: for patients still at risk at time $t$, the
probability that a patient who has the event by $t$ was assigned a
higher predicted risk than a patient who is event-free through $t$. It
ranges from 0.5 (no discrimination, equivalent to chance) to 1 (perfect
discrimination), on the same conceptual scale as Harrell's C-index, but
is evaluated separately at each fixed horizon (1, 3, 5 years here)
rather than averaged over the whole follow-up. 
  - The Brier score is the
time-dependent mean squared error between observed event status and
predicted event probability at $t$. It reflects both discrimination and
calibration jointly and ranges from 0 (perfect prediction) up to about
0.25 for an uninformative model at a 50% event rate, with lower values
indicating better overall accuracy. 
  - The Index of Prediction Accuracy
(IPA) rescales the Brier score relative to a reference model without
covariates (the marginal Kaplan-Meier estimate). IPA = 1 is perfect
prediction, IPA = 0 means the model is no better than the covariate-free
reference, and negative values mean the model performs *worse* than
that reference. 
The `Full model` estimates reported by `Score()` above
are already bootstrap-cross-validation corrected for optimism
(`split.method = "bootcv"`), analogous to the optimism-correction
applied to the C-index and calibration slope earlier.

As expected, AUC(t) decreases with a longer prediction horizon (1
year higher than 5 years) as predicting recurrence further into the
future is intrinsically harder, since more competing/intervening
information could in principle become available and the risk set
composition changes. AUC(t) is somewhat higher than the (time-averaged)
C-index above, which is an expected pattern. The two
target related but distinct estimands (Uno-type time-dependent AUC at a
fixed horizon vs. Harrell's C averaged across all comparable pairs over
the entire follow-up) and are not expected to numerically coincide. The
pattern is broadly similar to the multiple imputation-pooled AUC(t),
Brier score and IPA reported above. The complete-case and multiple
imputation-pooled estimates are not expected to coincide exactly,
given the different analysis populations (208 complete cases here vs.
all 248 patients) and the different treatment of missing data.

## A.10 : Conditional predicted survival curves {.unnumbered}

As hazard ratios are abstract, a profile-based illustration (predicted
disease-free survival (DFS) curves for a representative covariate pattern, contrasting
PMP22-high vs. PMP22-low with the other covariates held at their
reference (median/modal) values) makes the model concrete. It mirrors Figure 1C/1D of the original publication.

```{r}
#| label: profile-plot
#| fig-cap: "Model-based disease free survival predictions for PMP22-low vs. PMP22-high patients, other covariates fixed at reference (median/modal) values."

survplot(fit_cph, PMP22_f = c("low", "high"),
         conf.int = 0.95, conf = "bands",
         xlab = "Days since diagnosis",
         ylab = "Predicted disease-free survival probability",
         label.curves = TRUE)
```

## A.11 Reporting {.unnumbered}

<!--**No data-driven variable selection was performed at any stage of SAP
A.** All five predictors (`PMP22`, `pT`, `pN`, `G`, `ER`) were
pre-specified on clinical grounds (SAP A.1) and are retained in the
final model regardless of their individual statistical significance.
The only data-driven decisions made were the pre-specified functional-
form check for `pT`/`G` (A.4) and the pre-specified
proportional-hazards/interaction checks (A.7/A.8), neither of which are
variable-selection steps, since no predictor could be dropped by them.

The remaining TRIPOD reporting elements (A.11): patient flow, baseline
characteristics, the full pooled hazard-ratio table, and the
proportional-hazards/discrimination/calibration diagnostics are given
in A.2, A.5, A.8 and A.9 above. The limitations listed in the plan
(A.11) apply unchanged to this implementation.-->

To highlight how this pre-specified analysis plan would read once
written up for a clinical journal, this subsection presents a concise,
self-contained results write-up.

::: {.callout-tip collapse="true" title="Illustrative results write-up, as in a published paper"}

> **Methods (brief).** In a retrospective multicentre cohort of women
> with primary breast cancer [@tong2010gene], PMP22 gene expression and
> four established clinical factors (tumour size (pT), nodal status
> (pN), tumour grade (G) and estrogen receptor status (ER)) were
> evaluated as predictors of disease-free survival (DFS). All five
> predictors were specified a priori on clinical grounds; none was
> added or dropped based on its own association with outcome. The
> primary analysis fitted a multivariable Cox proportional-hazards
> model with all five predictors entered simultaneously, with missing
> predictor values handled by multiple imputation (20 imputed
> datasets, multivariate imputation by chained equations) and
> estimates combined with Rubin's rules. Because the proportional-
> hazards assumption was violated for ER (scaled Schoenfeld residuals),
> a pre-specified ER x log(follow-up time) interaction term was
> included to let the ER effect vary over follow-up. A complete-case
> analysis served as a sensitivity check. Model performance was
> assessed for the corresponding model without the ER x log(follow-up
> time) term (constant ER effect) and summarised by Harrell's C-index, a bootstrap-based calibration slope,
> Nagelkerke's explained-variation measure (pseudo-R^2^), and
> time-dependent AUC, Brier score and Index of Prediction Accuracy
> (IPA) at 1, 3 and 5 years, each corrected for optimism by bootstrap
> resampling repeated within every imputed dataset.

#### Patients {.unnumbered}

Of 249 patients in the original cohort, `r nrow(df)` were available in
the public analysis dataset. Of these, `r nrow(cc)`
(`r round(100*nrow(cc)/nrow(df),0)`%) had complete data on all five
predictors and were additionally used for complete-case sensitivity
analyses; the remaining `r nrow(df)-nrow(cc)` patients
(`r round(100*(nrow(df)-nrow(cc))/nrow(df),0)`%) had at least one
missing predictor and were included, via multiple imputation, in the
primary analysis population of all `r nrow(df)` patients (Figure 1).
Over a median potential follow-up of
`r round(summary(revkm)$table["median"]/365.25,1)` years (reverse
Kaplan-Meier estimate), `r sum(df$rezs)`
(`r round(100*mean(df$rezs),1)`%) of `r nrow(df)` patients experienced
disease recurrence or death.

```{r}
#| label: report-flow
#| fig-cap: "Figure 1. Patient flow."

grViz('
digraph report_flow {
  graph [layout = dot, rankdir = TB, nodesep = 0.4, ranksep = 0.45,
         fontsize = 11, fontname = Helvetica]
  node [shape = box, style = "rounded,filled", fillcolor = "#F2F2F2",
        color = "#4D4D4D", fontname = Helvetica, fontsize = 11,
        margin = "0.25,0.15"]
  edge [fontname = Helvetica, fontsize = 9, color = "#4D4D4D"]

  orig [label = "249 patients enrolled\\n(original cohort)"]
  pub  [label = "248 patients in the\\npublic analysis dataset"]
  cc   [label = "208 complete cases"]

  orig -> pub [label = "1 patient not available\\nin public dataset"]
  pub  -> cc [label = "40 patients with\\n>=1 missing predictor"]
}
')
```

#### Baseline characteristics {.unnumbered}

Table 1 summarises the baseline distribution of all five predictors
among all `r nrow(df)` patients in the primary analysis population with 
missing values shown as their own category.

```{r}
#| label: report-table1
#| tbl-cap: "Table 1. Baseline characteristics of all analysed patients."

report_row <- function(varname, label, data = df) {
  x <- data[[varname]]
  cats   <- c(levels(x), "Missing")
  counts <- c(table(x, useNA = "no"), sum(is.na(x)))
  data.frame(Variable = c(label, rep("", length(cats) - 1)),
             Category = cats,
             `N (%)`  = sprintf("%d (%.1f%%)", counts, 100 * counts / nrow(data)),
             check.names = FALSE)
}

report_tab1 <- rbind(
  report_row("PMP22_f", "PMP22 expression"),
  report_row("pN_f",    "Nodal status"),
  report_row("ER_f",    "Estrogen receptor status"),
  report_row("pT_f",    "Tumour stage"),
  report_row("G_f",     "Tumour grade")
)
kable(report_tab1, row.names = FALSE)
```

#### Results {.unnumbered}

The primary analysis was based on all 248 patients, with missing predictor
values handled by multiple imputation. All five prespecified variables were
retained in the multivariable model.

Because the proportional-hazards assumption was violated for ER status, the
ER effect was modelled as time-varying using an ER × log(follow-up time)
interaction. The effect of PMP22, tumour stage, nodal status and tumour
grade was modelled as proportional over follow-up.

```{r}
#| label: report-main-model-table
#| tbl-cap: "Table 2. Adjusted association between PMP22 expression and disease-free survival."

i_pmp22 <- match("PMP22 (high vs. low)", hr_table$term)

pmp22_report <- data.frame(
  `PMP22 expression` = "High vs. low",
  `Adjusted HR (95% CI)` =
    sprintf("%.2f (%.2f–%.2f)",
            hr_table$HR[i_pmp22],
            hr_table$CI_low[i_pmp22],
            hr_table$CI_high[i_pmp22]),
  `P value` =
    format.pval(hr_table$p_value[i_pmp22],
                digits = 3, eps = 0.001),
  check.names = FALSE
)

knitr::kable(
  pmp22_report,
  row.names = FALSE,
  align = c("l", "r", "r")
)

```

After adjustment for tumour stage, nodal status, tumour grade and the
time-varying effect of ER status, patients with high PMP22 expression had an
estimated 47% higher hazard of disease recurrence or death than patients
with low PMP22 expression (adjusted hazard ratio [HR] 1.47, 95% confidence
interval [CI] 0.94–2.29; p = 0.090; Table 2). The confidence interval is
compatible with effects ranging from a 6% lower hazard to more than twice
the hazard, indicating substantial uncertainty around the magnitude of the
association.

The complete-case sensitivity analysis gave a similar direction of
association, although the estimated effect was larger (HR 1.72, 95% CI
1.07–2.77; p = 0.025). The difference in estimates reflects, at least in
part, the additional information and uncertainty incorporated through
multiple imputation. (The complete-case analysis excluded the 40 patients
with missing predictor values.)

<!--
```{r}
#| label: report-er-time-varying
#| tbl-cap: "Time-specific hazard ratios for ER status from the model including an ER × log(follow-up time) interaction."

HR_ER_base <- hr_raw[
  hr_terms == "ER (positive vs. negative, at t=1 day)"
]

HR_ER_int <- hr_raw[
  hr_terms == "ER x log(time)"
]

HR_ER_1y <- HR_ER_base * HR_ER_int^log(365)
HR_ER_3y <- HR_ER_base * HR_ER_int^log(3 * 365)
HR_ER_5y <- HR_ER_base * HR_ER_int^log(5 * 365)

er_report <- data.frame(
  `Follow-up` = c("1 year", "3 years", "5 years"),
  `ER positive vs. negative, HR` =
    sprintf("%.2f", c(HR_ER_1y, HR_ER_3y, HR_ER_5y))
)

knitr::kable(
  er_report,
  row.names = FALSE,
  align = c("l", "r")
)

```

The effect of ER status varied over follow-up (ER × log[time] interaction,
p = 0.007). The estimated hazard ratio for ER-positive versus ER-negative disease was
0.24 at 1 year, 0.51 at 3 years and 0.73 at 5 years.-->

Figure 3 translates the estimated PMP22 effect into predicted disease-free
survival for patients with otherwise identical reference covariate values.
Predicted disease-free survival was lower for PMP22-high than for PMP22-low
patients, consistent with the adjusted HR above 1. The curves represent a
single covariate pattern and should not be interpreted as population-averaged survival estimates.


```{r}
#| label: report-survplot
#| fig-cap: "Figure 3. Model-based disease-free survival according to PMP22 expression, with other covariates fixed at reference values."

survplot(
  fit_cph,
  PMP22_f = c("low", "high"),
  conf.int = 0.95,
  conf = "bands",
  xlab = "Days since diagnosis",
  ylab = "Predicted disease-free survival probability",
  label.curves = TRUE
)

```

*Model performance*

Because the time-varying ER effect does not yield a single fixed prognostic score
over follow-up, the performance measures reported below refer to the
corresponding model with a constant ER effect. They should therefore be
interpreted as performance measures for this simplified prediction model,
not as direct validation of the primary time-varying-effect model.

```{r}
#| label: report-table3
#| tbl-cap: "Table 3. Optimism-corrected performance of the Cox model with a constant ER effect."

se_fmt <- function(x)
  ifelse(is.na(x), "–", sprintf("%.3f", x))

# Overall performance measures
perf_overall <- data.frame(
  Metric = c(
    "C-index",
    "Calibration slope"
  ),
  Horizon = "Overall",
  Estimate = sprintf(
    "%.3f",
    c(pooled_C, pooled_slope)
  ),
  SE = se_fmt(
    c(pooled_se, pooled_slope_se)
  )
)

# Time-dependent AUC
perf_auc <- data.frame(
  Metric = "Time-dependent AUC",
  Horizon = horizon,
  Estimate = sprintf(
    "%.3f",
    pooled_auc$estimate
  ),
  SE = se_fmt(
    pooled_auc$se
  )
)

# Combine and display
knitr::kable(
  rbind(perf_overall, perf_auc),
  row.names = FALSE,
  align = c("l", "l", "r", "r")
)
```

<!-- ALL MEASURES
```{r}
#| label: report-table32
#| tbl-cap: "Table 3. Optimism-corrected performance of the Cox model with a constant ER effect."

se_fmt <- function(x)
  ifelse(is.na(x), "–", sprintf("%.3f", x))

perf_overall <- data.frame(
  Metric = c(
    "C-index",
    "Calibration slope",
    "Nagelkerke pseudo-R²"
  ),
  Estimate = sprintf(
    "%.3f",
    c(pooled_C, pooled_slope, pooled_R2)
  ),
  SE = se_fmt(
    c(pooled_se, pooled_slope_se, pooled_R2_se)
  )
)

perf_td <- data.frame(
  Metric = rep(
    c("Time-dependent AUC",
      "Brier score",
      "Index of Prediction Accuracy"),
    each = 3
  ),
  Horizon = rep(horizon, 3),
  Estimate = sprintf(
    "%.3f",
    c(
      pooled_auc$estimate,
      pooled_brier$estimate,
      pooled_ipa$estimate
    )
  ),
  SE = se_fmt(
    c(
      pooled_auc$se,
      pooled_brier$se,
      rep(NA, 3)
    )
  )
)

knitr::kable(
  rbind(
    transform(perf_overall, Horizon = "Overall")[
      , c("Metric", "Horizon", "Estimate", "SE")
    ],
    perf_td
  ),
  row.names = FALSE,
  align = c("l", "l", "r", "r")
)

```
-->

The optimism-corrected C-index was 0.67 (SE 0.014), indicating moderate
discrimination. The calibration slope was 0.88 (SE 0.153). Time-dependent
AUCs were 0.66 at 1 year, 0.70 at 3 years and 0.67 at 5 years.

### Conclusion {.unnumbered}

In this cohort, higher PMP22 expression was associated with a higher hazard
of disease recurrence or death after adjustment for prespecified clinical
factors. The estimated effect was clinically meaningful in magnitude
(HR 1.47), but the confidence interval was wide and included the null,
indicating that the available data do not precisely determine the size of
the association.
:::

<!-- OLD VERSION
#### Prognostic model {.unnumbered}

Unadjusted Kaplan-Meier curves of disease-free survival by PMP22
category (complete cases) are shown in Figure 2 for descriptive purposes
only. As PMP22 and ER status are associated in this cohort, these crude
curves are confounded; they were not used to judge the prognostic value
of PMP22 or to select predictors, which rests on the adjusted model
below.

```{r}
#| label: report-km-by-pmp22
#| fig-cap: "Figure 2. Unadjusted Kaplan-Meier estimates of disease-free survival by PMP22 category, complete cases. Descriptive only; confounded by the PMP22-ER association."

fit_km_pmp22 <- survfit(Surv(days, rezs) ~ PMP22_f, data = cc)
ggsurvplot(fit_km_pmp22, data = cc, conf.int = TRUE, pval = FALSE,
           xlab = "Days since diagnosis",
           ylab = "Disease-free survival probability",
           legend.title = "PMP22", legend.labs = c("low", "high"))

```

```{r}
#| label: report-hr-prep

keep_rows <- hr_table$term != "ER x log(time)"
report_hr <- hr_table[keep_rows, c("term", "HR", "CI_low", "CI_high", "p_value")]
d_rep     <- hr_digits[keep_rows]
report_hr$`95% CI` <- paste0(fmt_hr(report_hr$CI_low, d_rep), "-",
                             fmt_hr(report_hr$CI_high, d_rep))
report_hr$HR       <- fmt_hr(report_hr$HR, d_rep)   # ER at t = 1 day with 4 decimals
report_hr$p_value  <- format.pval(report_hr$p_value, digits = 3, eps = 0.001)
report_hr <- report_hr[, c("term", "HR", "95% CI", "p_value")]

# derived time-specific ER HRs use the unrounded estimates
HR_ER_base <- hr_raw[hr_terms == "ER (positive vs. negative, at t=1 day)"]
HR_ER_int  <- hr_raw[hr_terms == "ER x log(time)"]
HR_ER_1y   <- HR_ER_base * HR_ER_int^log(365)
HR_ER_3y   <- HR_ER_base * HR_ER_int^log(3 * 365)
HR_ER_5y   <- HR_ER_base * HR_ER_int^log(5 * 365)

```
-->
<!--After adjustment for tumour size, nodal status, tumour grade and the time-dependent effect of ER
status, higher PMP22 expression was associated with an increased
hazard of recurrence or death (adjusted hazard ratio [HR]
`r hr_table$HR[hr_table$term=="PMP22 (high vs. low)"]`, 95% CI
`r hr_table$CI_low[hr_table$term=="PMP22 (high vs. low)"]`-`r hr_table$CI_high[hr_table$term=="PMP22 (high vs. low)"]`;
p = `r format.pval(hr_table$p_value[hr_table$term=="PMP22 (high vs. low)"], digits=3, eps=0.001)`;
Table 2, Figure 3). Full adjusted hazard ratios for all pre-specified
predictors are given in Table 2.-->
<!--
After adjustment for established clinical predictors (tumour size, nodal status, tumour grade and the time-dependent effect of ER
status), higher PMP22 expression
was associated with a higher hazard of disease recurrence or death (adjusted
HR `r hr_table$HR[hr_table$term=="PMP22 (high vs. low)"]`, 95% CI
`r hr_table$CI_low[hr_table$term=="PMP22 (high vs. low)"]`-`r hr_table$CI_high[hr_table$term=="PMP22 (high vs. low)"]`;
p = `r format.pval(hr_table$p_value[hr_table$term=="PMP22 (high vs. low)"], digits = 3, eps = 0.001)`).
This association was not statistically significant at the pre-specified
significance level of 0.05, and the confidence interval includes 1. Therefore, the data 
provide only suggestive, not conclusive, evidence for PMP22 as a prognostic
biomarker in primary breast cancer. 

```{r}
#| label: report-table2
#| tbl-cap: "Table 2. Multivariable-adjusted hazard ratios for disease-free survival (Cox proportional-hazards model; multiple imputation, m = 20, Rubin's rules)."

kable(report_hr, row.names = FALSE,
      col.names = c("Predictor", "Hazard ratio", "95% CI", "p-value"))

```

```{r}
#| label: report-table2B
#| tbl-cap: "Table 2. Multivariable-adjusted hazard ratios for disease-free survival (Cox proportional-hazards model; multiple imputation, m = 20, Rubin's rules). pT and G were modelled as ordinal quasi-continuous variables (HR per category/grade). ER status × time: ratio of the ER hazard ratios per one-unit increase in log(days)."

row_of <- function(term, label) {
  i  <- match(term, hr_table$term)
  d  <- hr_digits[i]
  c(label,
    fmt_hr(hr_table$HR[i], d),
    paste0(fmt_hr(hr_table$CI_low[i], d), "-", fmt_hr(hr_table$CI_high[i], d)),
    format.pval(hr_table$p_value[i], digits = 3, eps = 0.001))
}

tab2 <- rbind(
  c("**Main exposure**", "", "", ""),
  row_of("PMP22 (high vs. low)",          "PMP22 high vs. low"),
  c("**Adjustment variables**", "", "", ""),
  row_of("pT (per category)",             "pT"),
  row_of("pN (positive vs. negative)",    "pN node-positive vs. node-negative"),
  row_of("G (per grade)",                 "G"),
  row_of("ER x log(time)",                "ER status × time")
)

kable(tab2, row.names = FALSE,
      col.names = c("Variable", "HR", "95% CI", "P value"),
      align = c("l", "r", "r", "r"))

```

```{r}
#| label: report-table2C
#| tbl-cap: "Table 2. Multivariable-adjusted hazard ratios for disease-free survival (Cox proportional-hazards model; multiple imputation, m = 20, Rubin's rules). pT and G were modelled as ordinal quasi-continuous variables (HR per category/grade). The ER hazard ratio refers to 1 day after diagnosis (reference point of the interaction); for ER status × time, only the p-value of the ER × log(days) interaction term (time-dependent ER effect) is shown."

row_of <- function(term, label) {
  i  <- match(term, hr_table$term)
  d  <- hr_digits[i]
  c(label,
    fmt_hr(hr_table$HR[i], d),
    paste0(fmt_hr(hr_table$CI_low[i], d), "-", fmt_hr(hr_table$CI_high[i], d)),
    format.pval(hr_table$p_value[i], digits = 3, eps = 0.001))
}

# interaction term: p-value only
row_p_only <- function(term, label) {
  i <- match(term, hr_table$term)
  c(label, "", "", format.pval(hr_table$p_value[i], digits = 3, eps = 0.001))
}

tab2 <- rbind(
  c("**Main exposure**", "", "", ""),
  row_of("PMP22 (high vs. low)",                      "PMP22 high vs. low"),
  c("**Adjustment variables**", "", "", ""),
  row_of("pT (per category)",                         "pT"),
  row_of("pN (positive vs. negative)",                "pN node-positive vs. node-negative"),
  row_of("G (per grade)",                             "G"),
  row_of("ER (positive vs. negative, at t=1 day)",    "ER positive vs. negative (at t = 1 day)"),
  row_p_only("ER x log(time)",                        "ER status × time")
)

kable(tab2, row.names = FALSE,
      col.names = c("Variable", "HR", "95% CI", "P value"),
      align = c("l", "r", "r", "r"))

```

The ER effect was not constant over follow-up and should be interpreted as time-dependent rather than as a single
hazard ratio.The ER (positive
vs. negative) hazard ratio was `r round(HR_ER_1y,2)` at 1 year,
`r round(HR_ER_3y,2)` at 3 years, and `r round(HR_ER_5y,2)` at 5 years
(modelled via an ER x log[time] interaction; Table 2 reports the HR at
1 day after diagnosis, `r sprintf("%.4f", HR_ER_base)`, i.e., the reference
point of the interaction term). PMP22, tumour size, nodal status and tumour
grade were modelled as having proportional effects throughout
follow-up.

```{r}
#| label: report-survplot1
#| fig-cap: "Figure 3. Model-based disease-free survival predictions for PMP22-low vs. PMP22-high patients, other covariates fixed at reference (median/modal) values; complete cases."

survplot(fit_cph, PMP22_f = c("low", "high"),
         conf.int = 0.95, conf = "bands",
         xlab = "Days since diagnosis",
         ylab = "Predicted disease-free survival probability",
         label.curves = TRUE)

```

Figure 3 translates the adjusted PMP22 association into predicted
disease-free survival curves for a patient with reference covariate
values (median/modal); the predicted survival probability is lower for
PMP22-high than for PMP22-low patients, consistent with the adjusted
hazard ratio above 1. The curves are based on the complete-case
standard model with a constant ER effect, the shaded bands indicate
confidence intervals, and the curves refer to a single covariate
pattern and are not population-averaged.

#### Model performance {.unnumbered}

```{r}
#| label: report-table31
#| tbl-cap: "Table 3. Model performance of the standard (non-time-varying) model, multiple imputation-pooled (m = 20); bootstrap-corrected for optimism throughout."

se_fmt <- function(x) ifelse(is.na(x), "--", sprintf("%.3f", x))

perf_overall <- data.frame(
  Metric   = c("C-index (discrimination)", "Calibration slope",
               "Explained variation (Nagelkerke pseudo-R2)"),
  Horizon  = "--",
  Estimate = sprintf("%.3f", c(pooled_C, pooled_slope, pooled_R2)),
  SE       = se_fmt(c(pooled_se, pooled_slope_se, pooled_R2_se))
)

perf_td <- data.frame(
  Metric   = rep(c("AUC(t)", "Brier score", "Index of Prediction Accuracy"), each = 3),
  Horizon  = rep(horizon, 3),
  Estimate = sprintf("%.3f", c(pooled_auc$estimate, pooled_brier$estimate, pooled_ipa$estimate)),
  SE       = se_fmt(c(pooled_auc$se, pooled_brier$se, rep(NA, 3)))
)

kable(rbind(perf_overall, perf_td), row.names = FALSE,
      col.names = c("Metric", "Horizon", "Estimate", "SE"))
```

Discrimination (C-index `r round(pooled_C,2)`) and calibration
(bootstrap-corrected calibration slope `r round(pooled_slope,2)`) are
reported in Table 3; the explained-variation
measure (pseudo-R^2^ `r round(pooled_R2,2)`) indicates that a
substantial share of the variation in disease-free survival remains
unexplained by these five predictors, as is typical for this type of
model. Time-dependent AUC and IPA are generally expected to decline
somewhat at longer horizons (Table 3), as prediction becomes harder
further into the future. All measures in Table 3 refer to the same
standard proportional-hazards model with a constant ER effect, which
differs from the model in Table 2 by the omitted ER x log(time) term;
performance may therefore be slightly underestimated or otherwise
mis-estimated, in particular early or late in follow-up.

#### Interpretation {.unnumbered}

Higher PMP22 expression was independently associated with a higher
hazard of disease recurrence or death after adjustment for established
clinical predictors, supporting its candidacy as a prognostic
biomarker in primary breast cancer. The ER effect was not constant
over follow-up and should be interpreted as time-dependent rather than
as a single hazard ratio. Discrimination, calibration and prediction
accuracy are summarised in Table 3; as for most prognostic models based
on a handful of routinely available clinical and tumour-biological
predictors, imperfect performance is expected, and further external
validation would be needed before clinical use. The performance
measures refer to the model with a constant ER effect; the model with
time-varying ER effect was not validated separately.
:::
-->

------------------------------------------------------------------------

# SAP B: Univariable pre-screening. A common but flawed practice {#sec-analysis_uni}

## Overview {.unnumbered}

The workflow below is deliberately built to reproduce mistakes that are
 routinely seen in applied practice. It screens each candidate
predictor's univariable ( marginal) association with the outcome, keeps
only variables that pass a significance filter, and treats the resulting
reduced model as if it had been pre-specified.

## B.1 Univariable screening {.unnumbered}

Each predictor is tested **on its own**, using whatever cases happen to
be non-missing for that single variable. Therefore, the sample size silently differs from test to test, a detail that is rarely
reported in practice.

```{r}
#| label: planB-univariable

vars <- c("PMP22", "pT", "pN", "G", "ER")
uni_results <- do.call(rbind, lapply(vars, function(v) {
  d <- df[!is.na(df[[v]]), ]
  form <- as.formula(paste0("Surv(days, rezs) ~ ", v))
  fit <- coxph(form, data = d)
  s <- summary(fit)
  data.frame(variable = v, n = nrow(d),
             HR = round(s$coefficients[, "exp(coef)"], 2),
             p_value = s$coefficients[, "Pr(>|z|)"])  # full precision; used for the screen below
}))

uni_display <- uni_results
uni_display$p_value <- format.pval(uni_display$p_value, digits = 3, eps = 0.001)
kable(uni_display, row.names = FALSE,
      caption = "Step 1: univariable Cox regressions, each on its own available cases.")
```

## B.2 Apply a significance filter {.unnumbered}

Here we will use 0.05 as significance criterion. 

```{r}
#| label: planB-filter

alpha_screen <- 0.05
selected <- uni_results$variable[uni_results$p_value < alpha_screen]
dropped  <- setdiff(vars, selected)
cat("Retained after univariable screen (p <", alpha_screen, "):",
    paste(selected, collapse = ", "), "\n")
cat("Dropped (including, notably, PMP22 — the variable of primary interest):",
    paste(dropped, collapse = ", "), "\n")
```

PMP22 (p = `r format.pval(uni_results$p_value[uni_results$variable=="PMP22"], digits = 3, eps = 0.001)`) and `G` are
dropped outright, and `pT` (p =
`r format.pval(uni_results$p_value[uni_results$variable=="pT"], digits = 3, eps = 0.001)`) narrowly misses the
threshold too. The screen never gives PMP22 a chance to be assessed
jointly with the clinical factors; exactly the assessment the study
was designed to make. This happens because PMP22's univariable association with disease-free survival is
masked by its correlation with ER. Only *conditional* on ER (and the
other factors) does PMP22's prognostic value emerge.

## B.3 Multivariable model on the surviving subset {.unnumbered}

```{r}
#| label: planB-multivariable

cc_b <- df[complete.cases(df[, selected]), ]
cat("Complete-case n for the reduced model:", nrow(cc_b), "\n")

form_b <- as.formula(paste("Surv(days, rezs) ~", paste(selected, collapse = " + ")))
fit_b_full <- coxph(form_b, data = cc_b)
summary(fit_b_full)

```

A further backward-elimination step (here by AIC) is commonly layered on top
"to simplify the model further":

```{r}
#| label: planB-stepwise

fit_b_step <- stepAIC(fit_b_full, direction = "backward", trace = 1)
summary(fit_b_step)

```

## B.4 Missing data {.unnumbered}

No missing-data mechanism is examined. Each stage above uses
whichever complete cases happen to be available for the variables
involved at that stage (247/248 at the univariable stage for `PMP22` and
`ER`, 230 for `pT`, 219 for `pN`, 227 for `G`; `r nrow(cc_b)` for the final
reduced model). These inconsistent sample sizes that are easy to lose track
of and rarely reconciled in reports built this way.

## B.5 Interactions and assumptions — not checked {.unnumbered}

In a workflow like this, interaction terms and modeling assumptions
(proportional hazards, functional form, influential points) are
typically **not** examined once a "final" reduced model has been reached. We
deliberately do not run them here either, to reflect this common practice. Part A shows what they would have revealed. (Notably,
the `ER` proportional-hazards violation, which affects conclusions about `PMP22` specifically.)

## B.6 Reporting {.unnumbered}

The coefficients, hazard ratios, confidence intervals and p-values from
`fit_b_step` would typically be reported as if `pN` and `ER` had been
the pre-specified variables of interest, with language such as
"independent prognostic factors," and with no acknowledgment that they
were chosen because of, not despite, their own association with the
outcome. The resulting standard errors and p-values do not account for
the selection process and are anti-conservative.

## B.7 Apparent-only performance {.unnumbered}

```{r}
#| label: planB-performance

concordance(fit_b_step)
```

This apparent (in-sample, non-validated) concordance is reported
directly, with no bootstrap or cross-validation correction for optimism.

# Summarized consequences {#sec-analysis_contrast}

```{r}
#| label: planB-summary-table

comparison <- data.frame(
  Aspect = c("PMP22 in final model?", "Basis for inclusion", "Missing data handling",
             "Interactions checked?", "PH assumption checked?",
             "Inference validity", "Performance reported"),
  `Plan A (recommended)` = c(
    "Yes -- HR ~1.5-1.8, adjusted (complete-case p<0.05; multiple imputation-pooled primary p~0.09)",
    "Pre-specified a priori (clinical knowledge)",
    "Multiple imputation (+ complete case analysis sensitivity)",
    "Yes, pre-specified set, joint test",
    "Yes, ER violation found and corrected",
    "Valid (no selection on outcome)",
    "Bootstrap optimism-corrected"),
  `Plan B` = c(
    "No, eliminated at univariable screen",
    "Univariable p-value < 0.05",
    "Ad hoc complete case, inconsistent n",
    "Not considered",
    "Not checked",
    "Invalid (ignores selection uncertainty)",
    "Apparent (in-sample) only"),
  check.names = FALSE
)
kable(comparison)

```
<br>

The two workflows applied to the **same 248 patients** reach opposite
conclusions about the variable the study was designed to evaluate. Plan
B is not a hypothetical strawman: the univariable p-values obtained here
(`PMP22` p $\approx$ 0.98, `pT` p $\approx$ 0.06, `G` p $\approx$ 0.70) closely match those
reported in the original publication's own univariable analysis. The authors explicitly *did not* use them for pre-screening, and it is
precisely because they used a pre-specified multivariable model instead that PMP22's prognostic value was discovered at all.

# Computing Environment

Display the computing environment for the analysis.

```{r}
#| label: sessionInfo
#| echo: FALSE
#| warning: FALSE

sessioninfo::session_info()

```
------------------------------------------------------------------------

**Use of AI:** The code implementing the two statistical analysis plans was generated with AI assistance (Claude, Anthropic). The authors critically reviewed, revised, and adapted AI-generated content and code.
